---
title: "Assessment"
subtitle: "Investigating Trends in Continuous Glucose Monitor Presecriptions in Scotland"
output: 
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show 
    theme: cosmo 
    css: styles.css
date: "2025-10-31"
---
# Introduction


# Methods

```{r setup, include = TRUE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r packages, results=FALSE}
#importing required packages
library(tidyverse)
library(janitor)
library(gt)
library(here)
library(sf) #draw geographical map
library(ggiraph) #interactive maps
```

```{r function-import-clean-names}
#creating function to import data then clean names
import_clean_names <- function(file_path){
  read_csv(file_path)%>% 
    clean_names()}
```

```{r import-hb-names}
#loading all health board names
hb_names_data <- import_clean_names("https://www.opendata.nhs.scot/dataset/9f942fdb-e59e-44f5-b534-d6e17229cc7b/resource/652ff726-e676-4a20-abda-435b98dd7bdc/download/hb14_hb19.csv") %>% select(hb, hb_name)
```

```{r hb-population-estimates}
#importing health board population estimate data
population_data_raw <- import_clean_names("https://www.opendata.nhs.scot/dataset/7f010430-6ce1-4813-b25c-f7f335bdc4dc/resource/27a72cc8-d6d8-430c-8b4f-3109a9ceadb1/download/hb2019_pop_est_03102025.csv")
```

```{r population-data-adding-discrete_age-groups}
#tidying health board population data and splitting ages into distinct groups
population_data <- population_data_raw %>% 
  filter(sex == "All", year >= 2017, hb != "S92000003") %>% #removing uneeded data
  mutate(age_under_16 = rowSums(.[,7:22]), # splitting ages into groups
         age_16_to_24 = rowSums(.[,23:31]),
         age_25_to_34 = rowSums(.[,32:41]),
         age_35_to_54 = rowSums(.[,42:61]),
         age_55_to_64 = rowSums(.[,62:71]),
         age_65_plus = rowSums(.[,72:97])) %>%
  select(year,hb,sex,all_ages,age_under_16:age_65_plus) #keeping only needed columns
  
#copying year 2024 data for 2025 as there is none
population_data <- population_data %>% 
  filter(year == 2024) %>% 
  mutate(year = 2025) %>%
  rbind(population_data)
```
# REMOVE

```{r cgm_data}
#reading csv for CGM that has been made
data_cgm <- read_csv(here("data", "data_filtered", "data_cgm.csv"))
```

```{r hb-boundaries, results = FALSE}
#importing health boards geographical boundaries 
hb_map <- st_read(here("data", "SG_NHS_HealthBoards_2019", "SG_NHS_HealthBoards_2019.shp")) %>% 
  clean_names() %>% 
  mutate(hb_name = paste("NHS", hb_name)) #changing names to have NHS infront 
```

```{r simd-import}
#Importing SIMD data
simd_data <- import_clean_names("https://www.opendata.nhs.scot/dataset/78d41fa9-1a62-4f7b-9edb-3e8522a93378/resource/acade396-8430-4b34-895a-b3e757fa346e/download/simd2020v2_22062020.csv") %>% 
  rename(simd_decile = simd2020v2country_decile) %>% #changing name for easier coding
  select(data_zone, hb, simd_decile) #only keeping needed columns
```

```{r gp-practice-data}
#importing data about gp practice names, locations and populations
gp_practice_data <- import_clean_names("https://www.opendata.nhs.scot/dataset/f23655c3-6e23-4103-a511-a80d998adb90/resource/47557411-7eda-4278-9d6d-d26ed2ceab5a/download/practice_contact_details_20251001_o") %>% 
  select(practice_code:practice_list_size, hb, data_zone)#keeping only needed columns
```

```{r diabetes-healthboard-numbers, results = FALSE}
#importing data from file created for number of diabetic patients by health board
diabetes_by_healthboard <- import_clean_names(
  here("docs","diabetes_rates_by_healthboard.csv"))
```

```{r import-all-prescription-data}
# #Creating path for map function to import all health board data
# links <- read_csv(here("docs", "file_links.csv")) %>%
#   select(link) %>%
#   pull()
# 
# #Loading all prescription data from 10/2017 to 2025 when CGMs were presribed
# raw_prescription_data <- links %>%
#   map(read_csv) %>% 
#   list_rbind() %>%
#   clean_names()
```

```{r tidying-CGM-data}
#Combining health boards which changed codes boundary changes do not matter
# data_cgm <- raw_prescription_data %>%
#   filter(str_detect(bnf_item_code, "214800001")) %>%
#   mutate(hbt = recode(hbt, "S08000018" = "S08000029", "S08000021" = "S08000031", "S08000023" = "S08000032", "S08000027" = "S08000030"),
#          paid_date_month = parse_date_time(paid_date_month, "ym"),
#          year = as.numeric(year(paid_date_month)),#done to allow joining
#          hbt = coalesce(hbt2014, hbt)) %>%
#   select(-c(hbt2014, prescribed_type, class_of_preparation_code, dmd_code)) %>%
#   left_join(hb_names_data, join_by(hbt == hb))

data_cgm <- data_cgm %>% 
  mutate(hbt = recode(hbt, "S08000018" = "S08000029", "S08000021" = "S08000031", "S08000023" = "S08000032", "S08000027" = "S08000030"), 
         year = as.numeric(year(paid_date_month))) %>% 
  left_join(hb_names_data, join_by(hbt == hb))
```

# Results


```{r creating-function-summary}
summary_function <- function(data, grouping){data%>% 
    filter(!is.na(hb_name)) %>%
    group_by(across(grouping)) %>% 
    summarise(total_cgm = sum(paid_quantity)) %>%
    left_join(population_data, join_by(hbt == hb, year == year)) %>% 
    mutate(total_per_100k = (total_cgm/all_ages)*100000)}
```

## CGM presriptions over time by healthboard

```{r filtering-data-graph-hb-100k}
#Adding health board names and population estimates to the CGM dataset and then calculating rate per 100k population
cgm_prescription_by_year_data <- summary_function(data = data_cgm, grouping = c("paid_date_month", "hb_name", "hbt", "year"))
```

```{r graph-function}
#Function to create graph of year bs CGM prescription per 100k
#When passing in faceting variable must pit ~ in front of it
faceted_graph_function <- function(data, faceting, legend_text_size){
  data %>% 
    ggplot(aes(x = paid_date_month, y = total_per_100k)) +
    geom_line() +
    facet_wrap(enquo(faceting)) +
    theme_bw() +
    labs(x = "Year", y = "CGM Prescriptions (per 100k)") +
    theme(strip.text = element_text(size = legend_text_size)) +
    scale_x_date(limits = as.Date(c(NA, "2025-12-01")))
  }
```

```{r graph-hb-100k}
#creating graph of CGM prescription per 100k by health board overtime
cgm_prescription_by_year_graph <- 
  faceted_graph_function(cgm_prescription_by_year_data, ~hb_name, 7) +
  labs(title = "CGM prescriptions from November 2017 to August 2025",
       subtitle = "By health board in Scotland")

cgm_prescription_by_year_graph
```

The graph depicts all health boards increasing prescriptions in CGMs steadily overtime from November 2017 to August 2025. But there are geographic differences in the rate and total number of prescription between healthboards. NHS Dumfries and Galloway, Shetland and Western Isles prescribing more per 100,000 population in 2025 than any other health boards, and NHS Ayrshire and Arran, Tayside and Orkney having lower rates. 

## Interactive map of CGM prescriptions/number of diabetes patients per heathboard in the years 2018 and 2023
```{r filtering-data-map}
#Filtering data for CGM map by joining with needed databases and calculating totals
cgm_map_data <- summary_function(data = data_cgm, grouping = c("hb_name","year","hbt")) %>%
  filter(year == 2023 | year == 2018) %>% 
  left_join(diabetes_by_healthboard, join_by(hb_name, year == year)) %>% 
  mutate(cgm_per_diabetic_patient = (total_cgm/number_diabetic_patients)) 

cgm_map_data_2018_and_2023 <- cgm_map_data %>% 
  select(hb_name, year,cgm_per_diabetic_patient) %>% 
  pivot_wider(names_from = year, values_from = cgm_per_diabetic_patient, names_prefix = "cgm_per_diabetic_patient_") %>% 
  mutate(tooltip_text = paste(hb_name,"\n", "CGM prescriptions per diabetic patient 2018:", round(cgm_per_diabetic_patient_2018, 2),"\n", "CGM prescriptions per diabetic patient 2023:", round(cgm_per_diabetic_patient_2023, 2))) %>% 
  right_join(cgm_map_data, join_by(hb_name)) %>% 
  left_join(hb_map, join_by(hb_name)) 
```

```{r making-map}
#Making and styling map
# cgm_map_2018_and_2023 <- cgm_map_data_2018_and_2023 %>%
#   ggplot(aes(fill = cgm_per_diabetic_patient, geometry = geometry)) +
#   geom_sf_interactive(
#     aes(data_id = hb_name, 
#         tooltip = tooltip_text)) +
#   scale_fill_distiller(palette = 3) +
#   theme_void() +
#   facet_grid(~year) +
#   labs(title = "Interactive map showing prescription of CGMs \n per patient with diabetes mellitus by Scottish health boards",
#        subtitle = "In the years 2018 and 2023",
#        fill = "Number of CGMs \n prescribed per patient with \n diabetes mellitus per year")
# 
# girafe(ggobj = cgm_map_2018_and_2023)
```

## Table showing the proportion of people in each age group by board compared to number of CGM prescriptions in 2024

```{r table-filtering}
#Filtering data for table
age_table_data <- summary_function(data = data_cgm, grouping = c("hb_name","year","hbt")) %>% 
  mutate(across(age_under_16:age_65_plus, ~.x/all_ages))
```

```{r table-creation}
#Creating and styling table
age_table_data %>% 
  filter(year == 2024) %>% 
  ungroup() %>% 
  select(hb_name, year, age_under_16:total_per_100k) %>% 
  gt() %>% 
  fmt_percent(columns = age_under_16:age_65_plus, decimals = 2) %>% 
  cols_label(hb_name = "Health Board",
             year = "Year",
             total_per_100k = "Number of CGM prescriptions (per 100K)",
             age_under_16 = "Under 16",
             age_16_to_24 = "16 to 24",
             age_25_to_34 = "25 to 35",
             age_35_to_54 = "35 to 54",
             age_55_to_64 = "55 to 64",
             age_65_plus =  "65 and older") %>% 
  opt_stylize(style = 6, color = "blue") %>% 
  tab_header(title = "Number of CGM prescriptions by Scottish Healthboard compared to percercentage of people in each age group",
             subtitle = "In 2024") %>% 
  tab_spanner(label = "Age",
              columns = age_under_16:age_65_plus)
```

## CGM prescriptions by GP practise SIMD
```{r filtering-data-graph-cgm-by-simd}
#Combining GP data with SIMD and finding total number of patients per GP by SIMD
gp_population_by_simd <- gp_practice_data %>% 
  left_join(simd_data, join_by(data_zone)) %>% 
  group_by(simd_decile) %>% 
  summarise(total_gp_population = sum(practice_list_size))

#Combing patients per SIMD into CGM data and changing for graph
cgm_prescription_by_simd_data <- data_cgm %>% 
  left_join(gp_practice_data, join_by(gp_practice == practice_code)) %>% 
  left_join(simd_data, join_by(data_zone, hb)) %>% 
  filter(!is.na(gp_practice_name)) %>% 
  group_by(simd_decile, paid_date_month) %>% 
  summarise(total_cgm = sum(paid_quantity)) %>% 
  left_join(gp_population_by_simd, join_by(simd_decile)) %>% 
  mutate(total_per_100k = (total_cgm/total_gp_population)*100000)
```

```{r graph-creation-simd}
#Creating and styling graph for CGM prescriptions by SIMD 
cgm_prescription_by_simd_graph <- cgm_prescription_by_simd_data %>% 
  mutate(simd_decile = factor(paste("Decile", simd_decile), 
                              levels = paste("Decile", 1:10))) %>% 
  faceted_graph_function(., ~simd_decile, 10) +
  labs(title = "CGM prescriptions from November 2017 to August 2025",
       subtitle = "By SIMD in Scotland")

cgm_prescription_by_simd_graph
```

# Limitations

# Conclusions
 
# References 




 

